---
title: "example_prep_non-spatial.rmd"
output: html_document
---

- Process (use two BHI examples: one non-spatial and one spatial) 
  - load data
  - change to long format and save as .csv 
  - explore raw data and see whether it makes sense
  - plot and describe 
  - explore status and trend calcluations (then move to functions.r)
  - save in layers folder and register data layers in layers.csv

## Intro: 
This section you can introduce the goal, what types of information or data are needed, data sources, the goal model, and how to approach trend calculation. We'll use Artisanal Fishing Opportunities (modified from [OHI-Baltic](https://github.com/OHI-Science/bhi/blob/a306fd0b596fd32a42eae755590dc60f48be7acc/baltic2015/prep/ICO/ico_prep.rmd)) as an example: 

### Background

_What this goal is trying to measure, what it means in your local context, etc_

This goal has three sub-components: stock, access, and need. For BHI we focus first on the stock sub-component and will use this as a proxy for the entire goal initially


### Goal model

_Record what the goal model and reference point, how to approach trend calculations, etc._ 

This goal has three sub-components: stock, access, and need. For BHI we focus first on the stock sub-component and will use this as a proxy for the entire goal initially

#### Status

Xao = Mean Stock Indicator Value / Reference pt
Stock indicators = two HELCOM core indicators assessed for good environemental status (each scored between 0 and 1 by BHI)
Reference pt = maximum possible good environmental status (value=1)

#### Trend

CPUE time series are available for all stations used for the HELCOM coastal fish populations core indicators. These data were provided by Jens Olsson (FISH PRO II project). To calculate GES status, full time series were used. Therefore, only one status time point and cannot calculate trend of status over time. Instead, follow approach from Bergström et al 2016, but only focus on the final time period for the slope (2004-2013).

Bergstrom et al. 2016. Long term changes in the status of coastal fish in the Baltic Sea. Estuarin, Coast and Shelf Science. 169:74-84

Method: 

1. Select final time period of trend assessment (2004-2013)
2. Use time series from both indicators, Key Species and Functional groups. For functional groups,include both cyprinid and piscivore time series.
3. For each time series: square-root transform data, z-score, fit linear regression, extract slope
4. Within each time series group (key species, cyprinid, piscivore), take the mean slope for each group within each basin
5. Within each basin take a mean functional group indicator slope (mean of cyprinid mean and piscivore mean)
6. For each basin take overall mean slope - mean of key species and functional group
7. Apply trend value for basin to all BHI regions (except in Gulf of Finland, do not apply Finnish site value to Estonia and Russian regions.)


### Data sources

_Where the data comes from, where it's stored, potential concerns with the data, why you included or excluded certain data, etc:_

HELCOM Core Indicator Abundance of coastal fish key functional groups (link...)

HELCOM Core Indicator Abundance of key coastal fish species (link...)

Good Environmental Status (GES) is assessed as either GES or sub-GES based on data times series using either a baseline or a trend approach, see explanation. There is only a single assessment for each region.

status qualifying comments: for one dataset if a monitoring station receives a "sub-GES" assessment, it is given a qualifier as "low" or "high".

Environmental status assessments provided by Jens Olsson (SLU) on June 18, 2016. See HELCOM FISH-PRO II (link...)

Data are housed in the folder 'AO/raw_data'

_Other comments you could also record here:_

Pros and cons of using these data: 
- Pros: ... 
- Cons: ... 

Reasons for exlcuding certain datasets... 


### Data prep process

#### setup 

This section will set up directories, functions, call commonly used libraries, etc, to prepare for the next steps of data prep. 

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

## source common libraries, directories, functions, etc
source('~/github/bhi/baltic2015/prep/common.r')
dir_ao    = file.path(dir_prep, 'AO')

## add a README.md to the prep directory with the rawgit.com url for viewing on GitHub
create_readme(dir_secchi, 'secchi_prep.rmd')

## rprojroot
root <- rprojroot::is_rstudio_project

## make_path() function to 
make_path <- function(...) rprojroot::find_root_file(..., criterion = is_rstudio_project)

dir_layers = make_path('baltic2015/layers') # replaces file.path(dir_baltic, 'layers') ???? what about dir_layers in common.r????

```

#### Read in data 

```{r read in data, echo = FALSE}



```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
