---
title: "example_prep.rmd"
output: html_document
---

## Introduction of OHI data prep

This document will help you get started on preparing and examing your data before calculating the OHI scores... We'll start with introduction to the R packages we use most frequently for data wrangling, resources for spatial analysis, and how to use Rmd files to present the data preparation process in a human readable manner. 

## R resources

how to use .rmd; packages that we use often, including data wrangling and spatial analysis

### Basic R commands

If you are learning R or need a refresher on basic commands: 
  - [hands-on basic R tutorial for R beginners](http://tryr.codeschool.com/levels/1/challenges/1)
  - [basic R cheatsheet](https://www.rstudio.com/wp-content/uploads/2016/06/r-cheat-sheet.pdf)
  
Packages we rely on the most for data wrangling: 
  - data wrangling: dplyr and tidyr tutorials 
    - [cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
    
Sometimes the data you need (eg. coverage of a carbon sequestering habitat) is contained in a spatial file. R is equipped to process, extract, and manipulate spatial data, similar to ArcGIS. Here are tutorials that cover most of the commands you will need: 

- [dplyr and tidry tutorial](http://jules32.github.io/2016-07-12-Oxford/dplyr_tidyr/)
- [how to use rmd](https://rawgit.com/eco-data-science/rmarkdown_R/master/1_rmarkdown.html)

- Process (use two BHI examples: one non-spatial and one spatial) 
  - load data
  - change to long format and save as .csv 
  - explore raw data and see whether it makes sense
  - plot and describe 
  - explore status and trend calcluations (then move to functions.r)
  - save in layers folder and register data layers in layers.csv



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
